WasteAndMaterialFootprint: A python package to quantify supply chain flows of waste and material in LCA databases

### 2024-01-30 latex_files/discussion.tex
update-manuscript

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
\documentclass[final,5p,authoryear]{elsarticle}
%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}
\usepackage{lineno}
\usepackage{booktabs}
% ---------------------------------------------------------------------
%             FOR DRAFTING / REVIEW (DELETE BEFORE SUBMISSION)
% ---------------------------------------------------------------------
\usepackage{caption}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage[tt=false]{libertine}
\usepackage{xcolor}
\definecolor{blue}{RGB}{33,150,209}
\definecolor{red}{RGB}{255,0,0}
\usepackage{xurl}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{pdfpages}
\usepackage{enumitem}
\usepackage{lineno}
\setlength{\linenumbersep}{5pt}
% Save the old \texttt command
\let\oldtexttt\texttt
% Redefine the \UrlFont to use a smaller typewriter font
\renewcommand{\UrlFont}{\ttfamily\small}
% Redefine the \texttt command to use a smaller font size and allow line breaks
\renewcommand{\texttt}[1]{{\ttfamily\small\nolinkurl{#1}}}

\usepackage{tcolorbox}
\newcommand{\cbox}[1]{
    \begin{tcolorbox}[hbox, colback=red!5!white, colframe=red!65!black, boxrule=0.25pt, boxsep=2pt, left=2pt, right=2pt, top=1pt, bottom=1pt]
        \small\sffamily #1
    \end{tcolorbox}
}

% ------------------------------------------------------

\journal{Resources, Conservation and Recycling}

\begin{document}

\linenumbers

    % SPOTLIGHTS (5 bullet points, max 120 characters each)
    {\Large Spotlights}
    \vspace{1em}
    \begin{description}[style=nextline]
        \item[Bullet 1: Critical context and background information on the problem addressed] Understanding waste and material flows is essential for developing the circular economy and supply chain resilience
        \item[Bullet 2: A brief overview of the key finding of the study (or findings if necessary)] A program was developed to quantify waste and material footprints of activities in life cycle assessment (LCA) databases
        \item[Bullet 3: The most radical, creative, disruptive or innovative aspect of the manuscript] Built to complement the Brightway LCA framework and ActivityBrowser, the tool is flexible and easy to use.
        \item[Bullet 4: The significance of the results to the environment, economics or society] The tool can be used to identify pre-consumer waste and material hotspots that are often hidden in supply chains.
        \item[Bullet 5: Future vision or the most important implications for continued research] Improved data availability and quality would enable more detailed and accurate waste and material footprinting.
    \end{description}

    \newpage
    % %% GRAPHICAL ABSTRACT
    % \begin{graphicalabstract}
    %     %\includegraphics{grabs}
    % \end{graphicalabstract}

    %% RESEARCH HIGHLIGHTS (3-5 bullet points, max 85 characters each)
    \begin{highlights}
        \item WasteAndMaterialFootprint, a new tool for quantifying waste and material flows in LCA.
        \item Assesses supply risks by calculating demand for critical materials.
        \item Simplifies quantification of user-specified waste and material categories.
        \item Rapidly identifies waste and material demand hotspots.
        \item Presents a case study of the battery supply chain.
    \end{highlights}

\begin{frontmatter}

    \title{WasteAndMaterialFootprint: A python package to quantify supply chain flows of waste and material in LCA databases}
    \author[1]{Stewart Charles McDowall\corref{cor1}}
    \ead{s.c.mcdowall@cml.leidenuniv.nl}
    \author[1]{Elizabeth Lanphere}
    \author[1]{Stefano Cucurachi}
    \author[1]{Carlos Felipe Blanco}

    \cortext[cor1]{Corresponding author:}

    \affiliation[1]{
        organization={Institute of Environmental Sciences (CML), Leiden University},
        addressline={P.O. Box 9518},
        city={Leiden},
        postcode={2300 RA},
        state={South Holland},
        country={The Netherlands}}

    \begin{abstract}
        The `circular economy' is a concept focused on the prevention of waste and the reuse of resources. Thus, identifying and quantifying waste and material flows are of fundamental importance. Life Cycle Assessment (LCA) is powerful for this end, given its capacity to pinpoint hotspots of environmental impact throughout the life cycle of products and services, those where the implementation of circular principles could be most effective.

        Introducing WasteAndMaterialFootprint, a Python tool extending the Brightway framework for flexible, user-defined quantification of supply-chain demands in current and future scenarios. This tool streamlines database manipulation for LCA practitioners and integrates methods to aggregate and analyse waste and material flows, facilitating rapid hotspot identification.

        A case study on battery supply chains demonstrates the tool's utility. WasteAndMaterialFootprint quantifies and compares demands and potential environmental impacts, aiding sustainable decision-making. It contributes to a more sustainable, circular economy by providing detailed material usage and waste generation analysis.

    \end{abstract}

    % KEYWORDS (max 6)
    \begin{keyword}
        circular economy \sep waste \sep material \sep life cycle assessment \sep critical raw material \sep supply chain
    \end{keyword}

\end{frontmatter}

%% main text
\section{Introduction}
\label{sec:introduction}
\input{introduction.tex}

\section{Methodology}
\label{sec:methodology}
\input{methodology.tex}

\section{Results}
\label{sec:results}
\input{results.tex}

\section{Discussion}
\label{sec:discussion}
\input{discussion.tex}
\cbox{Section word count: aim for 400?}
\begin{itemize}
    \item Summary of the results, 1 par
    \item -----------------------------
    \item Advantages of the WMF tool, 1 par
    \item -----------------------------
    \item Limitations, 1 par
    \item -----------------------------
    \item Future possibilities, 1 par
\end{itemize}

\section{Conclusions}
\cbox{Section word count: aim for 150?}

This paper extends the state of knowledge by exploring the relationship between various waste aggregation methods and environmental damage indicators, contributing to a deeper understanding of life cycle waste inventories and their association with supply-chain risk and potential environmental damage.

\label{sec:conclusions}

% Back matter
\section*{Data availability} 
All data used in this analysis are publicly available online under the noted sources. The WasteAndMaterialFootprint tool is installable via the Python Package Index (PyPI) and is available at \url{https://pypi.org/project/WasteAndMaterialFootprint}.
The full source code for the WasteAndMaterialFootprint tool is available at \url{https://www.github.com/Stew-McD/WasteAndMaterialFootprint}.
A user guide and comprehensive documentation are available at \url{https://wasteandmaterialfootprint.readthedocs.io}.

\section*{CRediT authorship contribution statement}
\cbox{Co-authors, please check this and change as necessary.}
\textbf{Stewart Charles McDowall:} Methodology, Software, Validation, Formal analysis, Investigation, Data curation, Writing - original draft, Writing - review \& editing, Visualization.
\textbf{Elizabeth Lanphere:} Conceptualization, Methodology, Software, Validation, Writing - review \& editing.
\textbf{Stefano Cucurachi:} Conceptualization, Writing - review \& editing, Supervision, Project administration, Funding acquisition.
\textbf{Carlos Felipe Blanco:} Conceptualization, Writing - review \& editing, Supervision, Project administration, Funding acquisition.

\subsection*{Alternative CRediT statement}

\includegraphics[width=\columnwidth, height=7cm, keepaspectratio]{credit.pdf}

\section*{Declaration of competing interest}
The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

\section*{Acknowledgements}
\cbox{Should I put FutuRaMa here?}

    Part of this research project was financially supported by the European Union's Horizon 2020 research and innovation programme under the grant agreement No. 101058522 (project FutuRaM --- futuram.eu). The authors would like to thank the reviewers for their valuable comments and suggestions.

\section*{Supplementary material}
    The supplementary material contains the following:
    \begin{enumerate}
        \item List of waste and material categories
        \item Example output of the WasteAndMaterialFootprint tool
        \item Extended results of the case study
    \end{enumerate}

\bibliographystyle{elsarticle-harv}
\bibliography{references.bib}

% \appendix

% \section{Example output}
% \label{app:example_output}

\end{document}

>>>>

\begin{itemize}
    \item Summary of the results, 1 par
    \item -----------------------------
    \item Advantages of the WMF tool, 1 par
    \item -----------------------------
    \item Limitations, 1 par
    \item -----------------------------
    \item Future possibilities, 1 par
\end{itemize}

### 2024-01-30 latex_files/introduction.tex
update-manuscript

\cbox{Section word count: ~1700}
\cbox{Needs work to make it flow better, also shortened}

@@

Thus, we were motivated to create a flexible and policy-driven approach to understanding a product’s demand on the waste management system both now and in the future to ensure that models of circularity and improvements in technology and policy for waste are included in LCA modeling.

@@

\subsection{Material Demand in LCA}
In the context of a mineral-hungry renewable energy transition and recent geo-political tensions, more attention is being paid to the security of supply of materials, especially those considered `critical raw materials' (CRMs) \citep{eu2023crmstudy,hool2023crm,mancini2013supplysecurity,jrc2023supplychain,hartley2024cepolitics,salviulo2021supplychain}. Given the increasing focus placed on this, as well as ideals of improving resource efficiency and developing the circular economy, it is essential to understand the material demands of products and processes. While LCA seeks to model the technosphere (the sum of all supply-chains), its focus is often on the environmental impacts of the system---the endpoints--- rather than the primary material flows themselves.

A relatively new method, termed the crustal scarcity indicator (CSI)~\citep{arvidsson2020csi}, was developed to assess longterm global scarcity of minerals in life cycle assessment (LCA). This method introduced crustal scarcity potentials (CSPs) measured in kg silicon equivalents per kg element, derived from crustal concentrations. CSPs, provided for 76 elements, reflect the long-term global elemental scarcity based on crustal concentration proxies. The CSI, calculated by multiplying CSPs with extracted masses, effectively gauges the impact of elemental extraction.

@@

A significant portion of a product's total waste is generated during earlier stages such as resource extraction, transportation, and manufacturing, often remaining 'invisible' in traditional life cycle assessment (LCA) practices~\citep{laurenti2016wastefootprint}. This oversight in measuring and communicating a cradle-to-grave product waste footprint (PWF) highlights a gap in circular economy indicators. Traditional LCA does not typically view waste as having environmental significance by itself, focusing instead on emissions and resource use resulting from waste treatment. The environmental significance of waste and its correlation with other indicators has been a subject of extensive research. For example, studies have shown that popular resource footprints can cover a significant portion of environmental impact variance in product rankings~\citep{steinmann2017resourcefootprints}. However, correlations between various environmental indicators are not always consistent, as seen with the carbon footprint, which often does not correlate with other impact assessment scores~\citep{laurent2012carbonfootprint}. The aggregation of waste in PWFs raises concerns among LCA experts, regarding the uncertainties introduced by aggregated measures, as well as the potential misrepresentation of environmental performance due to differences in waste types~\citep{chen2021methoduncertainty,huijbregts2010energyfootprint}.

@@

\cite{laurenti2023wastefootprint} developed a method to calculate the waste footprint of a product or service based on solving the demand vectors of the activities, also presenting simple measures to quantify waste hazardousness and circularity. In that study, it was shown that the waste footprint correlates well with other LCIA methods, particularly human health. The method presented, however, is limited in its scope and flexibility, is computationally intensive, difficult to use, is not easily reproducible, and suffers from errors due to double counting. The WMF tool presented in the current work, builds on the work of Laurenti et.al., providing a more flexible, transparent and user-friendly approach to quantifying waste flows in LCA. Moreover, the WMF tool is not limited to waste, but can be used to quantify any supply-chain flow, such as water, gas, and critical raw materials.

@@

To better assess waste and material flows in LCA, the authors have developed a Python program built on the Brightway2 framework and designed to track these exchanges by translating them into indicators and impact categories. In this study, we present the WasteAndMaterialFootprint tool that enables LCA practitioners to easily aggregate the mass and volume of any desired exchange, and to creation of flexible categories to differentiate between waste types and End-of-Life (EOL) handling using (in this case) the Ecoinvent 3.9 cutoff database.

While methods with similar aims exist, they either lack flexibility and specificity \citep{foen2021ecofactors} or are cumbersome to apply and suffer from errors due to multiple counting \citep{laurenti2023wastefootprint}.

>>>>

@@

Thus, we were motivated to create a flexible and policy-driven approach to understanding a product`s demand on the waste management system both now and in the future to ensure that models of circularity and improvements in technology and policy for waste are included in LCA modeling.

@@

\subsection{Material Demand in LCA}
In the context of a mineral-hungry renewable energy transition and recent geo-political tensions, more attention is being paid to the security of supply of materials, especially those considered `critical raw materials' (CRMs) \citep{eu2023crmstudy,hool2023crm,mancini2013supplysecurity,jrc2023supplychain,hartley2024cepolitics,salviulo2021supplychain}. Given the increasing focus placed on this, as well as ideals of improving resource efficiency and developing the circular economy, it is essential to understand the material demands of products and processes. While LCA seeks to model the technosphere (the sum of all supply chains), its focus is often on the environmental impacts of the system---the endpoints--- rather than the primary material flows themselves.

A relatively new method, termed the crustal scarcity indicator (CSI)~\citep{arvidsson2020csi}, was developed to assess long-term global scarcity of minerals in life cycle assessment (LCA). This method introduced crustal scarcity potentials (CSPs) measured in kg silicon equivalents per kg element, derived from crustal concentrations. CSPs, provided for 76 elements, reflect the long-term global elemental scarcity based on crustal concentration proxies. The CSI, calculated by multiplying CSPs with extracted masses, effectively gauges the impact of elemental extraction.

@@

In LCA, waste flows are not considered as fundamental biosphere exchanges, but rather as technosphere flows. Waste produced by an activity is transferred to a relevant waste treatment activity where it is accepted `burden-free' and transformed into a combination of emissions and other waste `products'~\citep{guinee2021wasteisnotaservice}. There can be several treatment steps in this pathway leading, ultimately, to a mass of material being deposited in a landfill. In this system of waste accounting, the impacts apportioned to the waste-producing activity are a sum of those incurred by the transport, treatment, and final disposal of the waste into terrestrial or aquatic environments. In particular, the extensive work of~\cite{doka2024publications} has contributed significantly to understanding the environmental impacts of waste treatment processes and the long-term impacts of disposal.

A significant portion of a product's total waste is generated during earlier stages such as resource extraction, transportation, and manufacturing, often remaining `invisible' in traditional life cycle assessment (LCA) practices~\citep{laurenti2016wastefootprint}. This oversight in measuring and communicating a cradle-to-grave product waste footprint (PWF) highlights a gap in circular economy indicators. Traditional LCA does not typically view waste as having environmental significance by itself, focusing instead on emissions and resource use resulting from waste treatment. The environmental significance of waste and its correlation with other indicators has been the subject of extensive research. For example, studies have shown that popular resource footprints can cover a significant portion of environmental impact variance in product rankings~\citep{steinmann2017resourcefootprints}. However, correlations between various environmental indicators are not always consistent, as seen with the carbon footprint, which often does not correlate with other impact assessment scores~\citep{laurent2012carbonfootprint}. The aggregation of waste in PWFs raises concerns among LCA experts, regarding the uncertainties introduced by aggregated measures, as well as the potential misrepresentation of environmental performance due to differences in waste types~\citep{chen2021methoduncertainty,huijbregts2010energyfootprint}.

@@

\cite{laurenti2023wastefootprint} developed a method to calculate the waste footprint of a product or service based on solving the demand vectors of the activities, also presenting simple measures to quantify waste hazardousness and circularity. In that study, it was shown that the waste footprint correlates well with other LCIA methods, particularly human health. The method presented, however, is limited in its scope and flexibility, is computationally intensive, difficult to use, is not easily reproducible, and suffers from errors due to double counting. The WMF tool presented in the current work provides a more flexible, transparent, and user-friendly approach to quantifying waste flows in LCA. Moreover, the WMF tool is not limited to waste but can be used to quantify any supply-chain flow, such as water, gas, and critical raw materials.

@@

To better assess waste and material flows in LCA, the authors have developed a Python program built on the Brightway2 framework and designed to track these exchanges by translating them into indicators and impact categories. In this study, we present the WasteAndMaterialFootprint (WMF) tool that enables LCA practitioners to easily aggregate the mass and volume of any desired exchange, and to creation of flexible categories to differentiate between waste types and End-of-Life (EOL) handling using (in this case) the Ecoinvent 3.9 cutoff database.

The purpose of the WMF tool is not to quantify the environmental impacts of waste treatment, but rather to quantify the waste flows themselves, even those that are finally consumed by waste treatment processes. It provides, thus, not an impact assessment in the traditional sense, but an accounting of the waste generated by a product or service, regardless of the end-of-life fate of these flows. By definition, the development of the `circular economy' necessitates the reduction and ultimate elimination of waste---though whether this objective is thermodynamically impossible has long been the subject of lively debate by~\cite{ayres1998recycling},~\cite{reuter2012recyclinglimits} and many others. In any case, waste avoidance is of critical importance and by quantifying and classifying the waste total attributed to a product or service in an LCA database, the WMF tool provides a practical means to identify hotspots and opportunities for waste reduction and efficiency.

While methods with similar aims exist, they either lack flexibility and specificity \citep{foen2021ecofactors} or can be cumbersome to apply and suffer from errors due to multiple counting \citep{laurenti2023wastefootprint}.

### 2024-01-30 latex_files/main.tex
update-manuscript

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
\documentclass[final,5p,authoryear]{elsarticle}
%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}
\usepackage{lineno}
\usepackage{booktabs}
% ---------------------------------------------------------------------
%             FOR DRAFTING / REVIEW (DELETE BEFORE SUBMISSION)
% ---------------------------------------------------------------------
\usepackage{caption}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage[tt=false]{libertine}
\usepackage{xcolor}
\definecolor{blue}{RGB}{33,150,209}
\definecolor{red}{RGB}{255,0,0}
\usepackage{xurl}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{pdfpages}
\usepackage{enumitem}
\usepackage{lineno}
\setlength{\linenumbersep}{5pt}
% Save the old \texttt command
\let\oldtexttt\texttt
% Redefine the \UrlFont to use a smaller typewriter font
\renewcommand{\UrlFont}{\ttfamily\small}
% Redefine the \texttt command to use a smaller font size and allow line breaks
\renewcommand{\texttt}[1]{{\ttfamily\small\nolinkurl{#1}}}

@@

    \begin{abstract}
        The `circular economy' is a concept focused on the prevention of waste and the reuse of resources. Thus, identifying and quantifying waste and material flows are of fundamental importance. Life Cycle Assessment (LCA) is powerful for this end, given its capacity to pinpoint hotspots of environmental impact throughout the life cycle of products and services, those where the implementation of circular principles could be most effective.

@@

\end{frontmatter}

%% main text
\section{Introduction}
\label{sec:introduction}
\input{introduction.tex}

\section{Methodology}
\label{sec:methodology}
\input{methodology.tex}

\section{Results}
\label{sec:results}
\input{results.tex}

\section{Discussion}
\label{sec:discussion}
\input{discussion.tex}
\cbox{Section word count: aim for 400?}
\begin{itemize}
    \item Summary of the results, 1 par
    \item -----------------------------
    \item Advantages of the WMF tool, 1 par
    \item -----------------------------
    \item Limitations, 1 par
    \item -----------------------------
    \item Future possibilities, 1 par
\end{itemize}

\section{Conclusions}
\cbox{Section word count: aim for 150?}

@@

\label{sec:conclusions}

@@

\section*{Acknowledgements}
\cbox{Should I put FutuRaMa here?}

    Part of this research project was financially supported by the European Union's Horizon 2020 research and innovation programme under the grant agreement No. 101058522 (project FutuRaM --- futuram.eu). The authors would like to thank the reviewers for their valuable comments and suggestions.

>>>>

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
\documentclass[review,3p,authoryear]{elsarticle}
%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}
\usepackage{lineno}
\usepackage{booktabs}
% ---------------------------------------------------------------------
%             FOR DRAFTING / REVIEW (DELETE BEFORE SUBMISSION)
% ---------------------------------------------------------------------
\usepackage{caption}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage[tt=false]{libertine}
\usepackage{xcolor}
\definecolor{blue}{RGB}{33,150,209}
\definecolor{red}{RGB}{255,0,0}
\usepackage{xurl}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{pdfpages}
\usepackage{enumitem}
\usepackage{lineno}
\usepackage{float}
\setlength{\linenumbersep}{5pt}
% Save the old \texttt command
\let\oldtexttt\texttt
% Redefine the \UrlFont to use a smaller typewriter font
\renewcommand{\UrlFont}{\ttfamily\small}
% Redefine the \texttt command to use a smaller font size and allow line breaks
\renewcommand{\texttt}[1]{{\ttfamily\small\nolinkurl{#1}}}

@@

    \begin{abstract}
        \cbox{Abstract word count: 150, Limit: 150}

        The `circular economy' is a concept focused on the prevention of waste and the reuse of resources. Thus, identifying and quantifying waste and material flows are of fundamental importance. Life Cycle Assessment (LCA) is powerful for this end, given its capacity to pinpoint hotspots of environmental impact throughout the life cycle of products and services, those where the implementation of circular principles could be most effective.

@@

\end{frontmatter}
\cbox{Total word count: ~6000, Limit: 5000 (can be easily condensed)}

%% main text
\section{Introduction}
\label{sec:introduction}
\cbox{Section word count: 1650}
\cbox{Needs work to make it flow better, also shortened}
\input{introduction.tex}

\section{Methodology}
\label{sec:methodology}
\cbox{Section word count: 2000}
\input{methodology.tex}

\section{Results}
\cbox{Section word count: 1400}
\label{sec:results}
\input{results.tex}

\section{Discussion}
\label{sec:discussion}
\input{discussion.tex}
\cbox{Section word count: aim for 400?}

\section{Conclusions}
\label{sec:conclusions}
\cbox{Section word count: 250}
We have written the WasteAndMaterialFootprint (WMF) tool, an extension to the brightway2 LCA framework that enables the user to calculate the waste and material footprints of a product or service in an LCA database. It explodes the database, identifies upstream waste and material exchanges, edits them and writes matching custom WMF methods. These exchanges become pseudo-biosphere flows and thus, the footprint can be calculated as with the existing LCIA methods. The WMF tool can be easily customised by the user to calculate the footprints of other supply-chain flows such as water, gas, and critical raw materials.

Given that both waste generation and material demand are often strongly associated with the environmental impacts of an activity, it is important that they are included in the LCA. While there are numerous examples of existing and proposed methods that attempt to provide endpoint LCIA scores through convoluted formulae or subjective weighting, there is little consensus on their application and their complexity and lack of transparency can make them difficult to use and interpret.

The WMF tool, in contrast, gives LCA practitioners a simple, flexible, and transparent way to calculate the supply chain waste and material footprints, delivering results in standard units as direct aggregations of the relevant demand inventories.

@@

@@

\section*{Acknowledgements}
    Part of this research project was financially supported by the European Union's Horizon 2020 research and innovation programme under the grant agreement No. 101058522 (project FutuRaM --- futuram.eu). The authors would like to thank the reviewers for their valuable comments and suggestions.
